---
title: "Differential Gene Expression Analysis"
output: html_notebook
---

# Setup

## Load libraries

```{r}
library(DESeq2)
library(pheatmap)
library(RColorBrewer)
```

## Read count data

```{r}
count_table <- read.csv('raw_counts.tsv', sep='\t', row.names=1)
head(count_table)
```

```{r}
dim(count_table)
```

## Read the sample information

```{r}
sample_info <- read.csv('design.tsv', sep='\t', row.names=1)
sample_info
```

```{r}
dim(sample_info)
```

## Set factor levels

```{r}
factors <- factor(sample_info$Group)
groups <- unique(sample_info$Group)

sample_info$Group
```

```{r}
groups
```

## We want the control group to be after tgf-beta

```{r}
groups <- rev(groups)
groups
```

```{r}
sample_info$Group <- factors
sample_info$Group
```

## Create DESeq object

```{r}
dds <- DESeqDataSetFromMatrix(countData=count_table, colData=sample_info, design=~Group)
```

## Set the reference for the Group factor

```{r}
dds$Group <- relevel(dds$Group, ref='control')
```

## Filter out low gene counts

Keep genes with at least N counts \>= 10, where N = size of smallest group

```{r}
keep <- rowSums(counts(dds)>=10) >= min(table(sample_info$Group))
dds <- dds[keep,]
```

## Perform statistical tests and get result

```{r}
dds <- DESeq(dds, test='Wald', sfType='poscount')

deseq_result <- results(dds)
deseq_result <- as.data.frame(deseq_result)
head(deseq_result)
```

```{r}
dim(deseq_result)
```

## Add 'GeneName' column, save to tsv file

```{r}
deseq_result$GeneName <- row.names(deseq_result)

deseq_result <- subset(deseq_result,
                       select=c('GeneName', 'padj', 'pvalue', 'lfcSE', 'stat',  'log2FoldChange', 'baseMean'))

head(deseq_result)
```

```{r}
write.table(deseq_result, file='deseq_result.all.tsv', row.names=F, sep='\t')
```

## Extract DE genes with padj \< 0.05 and log2foldchange \<= -1 or \>= 1

```{r}
deg <- subset(deseq_result, padj<0.05 & abs(log2FoldChange)>=1)
dim(deg)
```

```{r}
dim(deseq_result)
```

## Order by padj ascending

```{r}
deg <- deg[order(deg$padj),]
head(deg)
```

## Write to tsv file

```{r}
write.table(deg, file='deseq_deg.tsv', row.names=F, sep='\t')
```

# Gene Expression Data Visualization

## Plot dispersion estimates

```{r}
plotDispEsts(dds, main='GSE203159 Dispersion Estimates')
```

## Create histogram of p-values

```{r}
hist(deseq_result$padj, breaks=seq(0,1,length=21), col='grey', border='white', xlab='', ylab='', ylim=c(0,8000), main='GSE203159 Frequencies of padj-values')
```

## Volcano plot

### Set colors

```{r}
old.pal <- palette(c('#00BFFF', '#FF3030'))
```

### Set margin size

```{r}
par(mar=c(4,4,2,1), cex.main=1.5)
```

### Plot values

Add legend for up- and down-regulation

```{r}
plot(deseq_result$log2FoldChange, -log10(deseq_result$padj), main='tgf-beta vs control', xlab='log2FC', ylab='-log10(Padj)', pch=20, cex=0.5)

with(subset(deseq_result, padj<0.05 & abs(log2FoldChange)>=1), points(log2FoldChange, -log10(padj), pch=20, col=(sign(log2FoldChange)+3)/2, cex=1))

legend('bottomleft', title=paste('Padj<', 0.05, sep=''), legend=c('down', 'up'), pch=20, col=1:2)
```

## PCA plot

### Variance stabilizing transformation

```{r}
vsd <- vst(dds, blind=FALSE)
```

### Use transformed values to generate PCA plot

```{r}
plotPCA(vsd, intgroup=c('Group'))
```

*Note: I couldn't find a fix to make the PCA plot's y-axis longer. In the future, I could manually plot the points but for the sake of this project I will continue.*

## Heatmaps

Heatmap of log transformed normalized counts using the top 10 genes

```{r}
normalized_counts <- counts(dds, normalized=T)
head(normalized_counts)
```

```{r}
transformed_counts <- log2(normalized_counts + 1)
head(transformed_counts)
```

```{r}
top_hits <- row.names(deg[1:10, ])
top_hits
```

```{r}
top_hits <- transformed_counts[top_hits,]
head(top_hits)
```

```{r}
pheatmap(top_hits, cluster_rows=FALSE, cluster_cols=FALSE)
```
